\begin{abstract}
\noindent
Few equations in applied mathematics have shaped a market the way the Black--Scholes PDE has: a parabolic equation, built on the same diffusion structure as the heat equation, that underpins trillions of dollars of derivatives traded every day. Yet the very assumption that makes Black--Scholes tractable -- a single constant volatility -- is also the assumption that markets contradict every minute through implied-volatility smiles, skews, and term-structure effects. Stochastic-volatility extensions such as the Heston model close this gap in theory, but in practice they push classical grid-based solvers into a curse of dimensionality, and they turn calibration -- recovering latent model parameters from quoted prices -- into a notoriously ill-conditioned inverse problem. This project asks whether a single, modern deep-learning framework can address both sides at once.

We use Physics-Informed Neural Networks (PINNs) to embed the option-pricing PDE directly into the training loss of a neural network, so the learned price surface is constrained to satisfy the underlying financial model rather than merely interpolating data. We build the framework in three deliberate stages of increasing difficulty. We start with the textbook constant-volatility Black-Scholes model as a controlled benchmark where a closed-form price exists and every PINN choice can be audited against it. We then extend the same architecture to the three-dimensional Heston stochastic-volatility PDE, with its additional variance dimension and a mixed-derivative term encoding asset--variance correlation, and benchmark it against a semi-analytical Fourier-quadrature reference. Finally, we invert the framework in the inverse Heston problem. The structural Heston parameters $(\kappa,\theta,\xi,\rho)$ are themselves made trainable, reparameterised through softplus / $\tanh$ bijections, optimised jointly with the network weights under a two-stage Adam + L-BFGS schedule and a Feller penalty, and recovered from $200$ noisy synthetic market prices.

On Black--Scholes, the PINN reproduces the analytical call price at the at-the-money point to within $0.56\%$ ($V_{\mathrm{PINN}}=6.927$ vs.\ $V_{\mathrm{BS}}=6.889$, PDE residual $\sim 10^{-3}$, $12$\,min on a single GPU). On Heston, the same framework prices the three-dimensional surface to within $0.44\%$ of the Fourier benchmark ($6.807$ vs.\ $6.837$, $25$\,min). On inverse Heston, the calibrated network fits the observed prices essentially perfectly ($R^{2}\approx 0.999999$, median relative pricing error $\approx 0.02\%$), yet the recovered correlation has the \emph{wrong sign} ($\hat{\rho}=+0.994$ against true $\rho=-0.7$).

Together, the three stages tell a coherent and practically important story. PINNs are already strong, differentiable forward pricers across both Black--Scholes and Heston, delivering prices and Greeks from a single mesh-free model. But the inverse case exposes a deeper lesson for quantitative finance: a near-perfect price fit does not imply structural-parameter recovery, because $\rho$ enters the Heston PDE only through the mixed term and is identifiable only from the implied-volatility skew that $200$ sparsely placed prices cannot carry. The path from Black--Scholes to Heston to inverse Heston therefore traces the same arc that practitioners walk every day -- from elegant theory, to realistic dynamics, to the hard reality of calibration -- and clarifies exactly where deep PDE solvers help, and where economically motivated regularisation, denser strike--maturity grids, and implied-volatility loss formulations remain indispensable.
\end{abstract}



