\section{Methods}\label{section:methods}

This section describes the mathematical models, the neural-network function class, the automatic-differentiation procedure, and the collocation, training, and calibration strategies used in the project. The overarching goal is to approximate option-price surfaces $V(S,\tau)$ and $V(S,v,\tau)$ by training Physics-Informed Neural Networks (PINNs) \parencite{raissi2019physics,karniadakis2021physics,cuomo2022scientific} that simultaneously satisfy the governing pricing PDEs and respect terminal payoff and asset-price boundary conditions. The Black--Scholes model \parencite{black1973pricing,merton1973rational} is used first as a controlled benchmark with a closed-form solution, before extending the same framework to the higher-dimensional Heston stochastic-volatility model \parencite{heston1993closed} and finally to an inverse calibration problem in which selected Heston parameters are recovered jointly with the network weights from noisy synthetic data.

\subsection{Notation and conventions}\label{subsec:notation}

Since the project moves between two financial models and a neural-network parameterization, several Greek letters carry context-dependent meanings. To avoid ambiguity, the following conventions are used throughout. The neural-network weights and biases are collected into a single parameter vector $\boldsymbol{w}$, written in boldface, so that the symbol $\theta$ is reserved exclusively for the Heston long-run variance. Likewise, the Black--Scholes rate sensitivity (Rho) is denoted $\rho_r$ to distinguish it from the Heston Brownian correlation $\rho$. Loss-function components carry subscripts identifying their role (PDE residual, initial condition, boundary condition, market data, Feller penalty), and each is associated with a non-negative weight $\lambda_{\bullet}$. Collocation-point counts use the symbol $N$ with a subscript indicating the corresponding loss term. The PDE residual operator acting on the network output is denoted $\mathcal{R}_{\boldsymbol{w}}$. These conventions are summarized in \cref{tab:notation}.

\begin{table}[!htbp]
    \centering
    \small
    \setlength{\tabcolsep}{6pt}
    \renewcommand{\arraystretch}{1.15}
    \caption{Notation used throughout the Methods section.}
    \label{tab:notation}
    \begin{tabularx}{\linewidth}{@{} l X @{}}
        \toprule
        \textbf{Symbol} & \textbf{Meaning} \\
        \midrule

        \multicolumn{2}{@{}l}{\textbf{Financial state variables}} \\
        $S$              & Underlying asset price \\
        $v$              & Instantaneous variance (Heston) \\
        $t$              & Calendar time \\
        $\tau = T-t$     & Time to maturity \\
        $T$              & Option maturity \\
        $K$              & Strike price \\
        $r$              & Risk-free interest rate \\

        \addlinespace[0.25em]
        \multicolumn{2}{@{}l}{\textbf{Model parameters}} \\
        $\sigma$         & Volatility (Black--Scholes) \\
        $\kappa$         & Mean-reversion speed (Heston) \\
        $\theta$         & Long-run variance (Heston) \\
        $\xi$            & Volatility of variance (Heston) \\
        $\rho$           & Brownian correlation (Heston) \\
        $v_0$            & Initial variance (Heston) \\

        \addlinespace[0.25em]
        \multicolumn{2}{@{}l}{\textbf{Pricing function and network}} \\
        $V$                       & Option price \\
        $V_{\boldsymbol{w}}$      & PINN approximation of $V$ \\
        $\boldsymbol{w}$          & Trainable network parameters (weights and biases) \\
        $\mathcal{R}_{\boldsymbol{w}}$ & PDE residual operator applied to $V_{\boldsymbol{w}}$ \\

        \addlinespace[0.25em]
        \multicolumn{2}{@{}l}{\textbf{Loss components and weights}} \\
        $\mathcal{L}_{\mathrm{pde}}$    & PDE-residual loss \\
        $\mathcal{L}_{\mathrm{ic}}$     & Initial-condition (terminal-payoff) loss \\
        $\mathcal{L}_{\mathrm{bc}}$     & Boundary-condition loss \\
        $\mathcal{L}_{\mathrm{data}}$   & Market-data loss (inverse problem) \\
        $\mathcal{L}_{\mathrm{Feller}}$ & Feller-condition penalty (inverse problem) \\
        $\lambda_{\bullet}$             & Non-negative weight of the corresponding loss term \\

        \addlinespace[0.25em]
        \multicolumn{2}{@{}l}{\textbf{Collocation point counts}} \\
        $N_f$            & Interior (PDE) collocation points \\
        $N_{\mathrm{ic}}$ & Initial-condition points \\
        $N_{\mathrm{bc}}$ & Boundary-condition points \\
        $N_m$            & Market-data points (inverse problem) \\

        \addlinespace[0.25em]
        \multicolumn{2}{@{}l}{\textbf{Greeks notation}} \\
        $\rho_r$         & Black--Scholes Rho, $\partial V/\partial r$ (distinct from Heston $\rho$) \\

        \bottomrule
    \end{tabularx}
\end{table}

\subsection{From option-pricing PDEs to diffusion equations}

The connection between option pricing and the physics of diffusion arises through the parabolic structure of the Black--Scholes equation. Under the standard geometric Brownian motion model,
\begin{equation}
    dS_t = \mu S_t\,dt + \sigma S_t\,dW_t,
\end{equation}
the value $V(S,t)$ of a European option satisfies, under risk-neutral pricing, the Black--Scholes PDE \parencite{black1973pricing,merton1973rational,shreve2004stochastic}
\begin{equation}\label{eq:bs_pde}
    \frac{\partial V}{\partial t}
    + \frac{1}{2}\sigma^2 S^2 \frac{\partial^2 V}{\partial S^2}
    + rS\frac{\partial V}{\partial S}
    - rV = 0,
\end{equation}
where $S$ is the underlying asset price, $t$ calendar time, $r$ the risk-free rate, and $\sigma$ the constant volatility \parencite{capinski2012blackscholes,profeta2010option}.

A key analytical insight is that \cref{eq:bs_pde} can be reduced to a standard diffusion equation through a change of variables \parencite{capinski2012blackscholes}. Introducing the log-moneyness coordinate and time-to-maturity variable
\begin{equation}
    x = \ln\!\left(\frac{S}{K}\right),
    \qquad
    \tau = T-t,
\end{equation}
together with a suitable exponential rescaling of the dependent variable, transforms \cref{eq:bs_pde} into the canonical heat equation
\begin{equation}
    \frac{\partial u}{\partial \tau}
    =
    \frac{\partial^2 u}{\partial x^2}.
\end{equation}
This diffusive structure motivates the PINN formulation: instead of solving the pricing PDE on a fixed mesh, a neural network is trained as a smooth global ansatz whose first and second derivatives can be obtained by automatic differentiation and substituted directly into the residual at sampled collocation points \parencite{raissi2019physics,baydin2018automatic}.

\subsection{The Black--Scholes model}

The Black--Scholes model \parencite{black1973pricing,merton1973rational} provides a closed-form benchmark for European option prices under the assumptions of constant volatility, continuous trading, frictionless markets, and lognormally distributed asset prices; standard stochastic-calculus treatments are given in \textcite{shreve2004stochastic,capinski2012blackscholes}. For a European call or put with strike $K$ and maturity $T$, define
\begin{equation}
    d_1 =
    \frac{\ln(S/K) + \left(r+\tfrac{1}{2}\sigma^2\right)T}
    {\sigma\sqrt{T}},
    \qquad
    d_2 = d_1 - \sigma\sqrt{T}.
\end{equation}
The corresponding call and put prices are
\begin{align}
    C &= S N(d_1) - K e^{-rT} N(d_2), \\
    P &= K e^{-rT} N(-d_2) - S N(-d_1),
\end{align}
where $N(\cdot)$ is the standard normal cumulative distribution function. These analytical prices are used as ground truth for validating the Black--Scholes PINN.

For the PINN implementation, the problem is recast in terms of time to maturity $\tau = T-t$. The terminal payoff at expiry therefore becomes an initial condition at $\tau=0$:
\begin{equation}
    V(S,0) = \max(S-K,0)
\end{equation}
for a European call. The asset-price boundaries are chosen to reflect the asymptotic behavior of the call,
\begin{align}
    V(0,\tau) &= 0, \\
    V(S_{\max},\tau) &\approx S_{\max} - K e^{-r\tau},
\end{align}
on the truncated domain $S \in [0, S_{\max}]$ with $S_{\max} = 300$ and $\tau \in [0, T_{\max}]$ with $T_{\max} = 1$.

\subsection{The Heston stochastic-volatility model}

The Black--Scholes assumption of constant volatility is mathematically convenient but empirically restrictive: market option prices typically exhibit volatility smiles and skews, suggesting that volatility itself should be modeled as stochastic \parencite{fouque2011multiscale,gulisashvili2012analytically,swishchuk2008time}. The Heston model \parencite{heston1993closed} extends the Black--Scholes framework with a square-root variance process of Cox--Ingersoll--Ross type \parencite{cox1985cir},
\begin{align}
    dS_t &= rS_t\,dt + \sqrt{v_t}\,S_t\,dW_t^S, \\
    dv_t &= \kappa(\theta-v_t)\,dt + \xi\sqrt{v_t}\,dW_t^v,
\end{align}
with instantaneous correlation $dW_t^S\,dW_t^v = \rho\,dt$. The parameters $\kappa>0$, $\theta>0$, $\xi>0$, $\rho\in(-1,1)$, and $v_0>0$ denote the mean-reversion speed, long-run variance, volatility of variance, Brownian correlation, and initial variance, respectively. The Feller condition $2\kappa\theta > \xi^2$ \parencite{feller1951two} ensures that the variance process does not reach zero with positive probability and is satisfied throughout this work by construction.

Under Heston dynamics, the option price depends on three variables, $V=V(S,v,t)$, and satisfies the two-dimensional pricing PDE \parencite{heston1993closed,ekstrom2010bs}
\begin{align}\label{eq:heston_pde}
    \frac{\partial V}{\partial t}
    &+ \tfrac{1}{2}vS^2 \frac{\partial^2 V}{\partial S^2}
    + \rho\xi vS \frac{\partial^2 V}{\partial S \partial v}
    + \tfrac{1}{2}\xi^2 v \frac{\partial^2 V}{\partial v^2} \notag \\
    &+ rS\frac{\partial V}{\partial S}
    + \kappa(\theta-v)\frac{\partial V}{\partial v}
    - rV = 0.
\end{align}
Compared with \cref{eq:bs_pde}, the Heston equation introduces an additional spatial dimension and a mixed derivative $\partial^2 V/\partial S\partial v$. Grid-based methods are correspondingly more demanding, since the computational cost grows rapidly as state variables are added. The PINN approach mitigates this curse of dimensionality through mesh-free Monte Carlo--style sampling of the interior domain \parencite{raissi2019physics,cuomo2022scientific}. More recent ``rough'' volatility models driven by fractional Brownian motion \parencite{gatheral2018volatility,bayer2016rough,euch2019roughheston,euch2018perfect} lie outside the scope of this project.

\paragraph{Semi-analytical benchmark.} The Heston PDE admits no elementary closed-form solution, but \textcite{heston1993closed} derived a semi-analytical pricing formula based on the characteristic function of the log-price. The European call is expressed as
\begin{equation}\label{eq:heston_cf_price}
    C(S_0,v_0,\tau) = S_0 P_1 - K e^{-r\tau} P_2,
\end{equation}
where the in-the-money probabilities $P_j$ are recovered by Fourier inversion,
\begin{equation}\label{eq:heston_pj}
    P_j = \tfrac{1}{2} + \tfrac{1}{\pi}\int_{0}^{\infty}\,\mathrm{Re}\!\left[\frac{e^{-i\phi\ln K}\,f_j(\phi;S_0,v_0,\tau)}{i\phi}\right]\!d\phi,\qquad j=1,2,
\end{equation}
and $f_j$ is the Heston characteristic function of \textcite{heston1993closed}. In this project, equations \eqref{eq:heston_cf_price}--\eqref{eq:heston_pj} are evaluated by direct numerical integration of each $P_j$ via the adaptive Gauss--Kronrod quadrature routine \texttt{scipy.integrate.quad} on the truncated frequency interval $[10^{-4},\,100]$, with the integrand expressed in the original (non-discounted) characteristic form. For maturities below $\tau \le 10^{-4}$ the payoff value is returned directly to avoid the singularity in $1/(i\phi)$. This semi-analytical surface is then used as ground truth for validating the Heston PINN. Related Fourier-pricing approaches include the FFT-based scheme of \textcite{carr1999fft} and the surveys in \parencite{albrecher2007heston,andersen2008simple}. The Fourier-cosine (COS) method of \textcite{fang2008cos} is a faster alternative that exploits the same characteristic function, but is not used here because the simpler quadrature implementation is sufficient for the moderate grid sizes considered.

\subsection{Theoretical foundations}\label{subsec:theory}

The PINN approach rests on two well-established results from the analysis of neural networks and modern scientific computing.

\paragraph{Universal approximation.} Feed-forward neural networks with at least one hidden layer can approximate any continuous function on a compact set to arbitrary precision, provided the activation function is non-polynomial and satisfies mild regularity conditions. \textcite{cybenko1989approximation} established this result for sigmoidal activations and \textcite{hornik1989multilayer} extended it to the multilayer case. Universal approximation is necessary but not sufficient for PINNs: it guarantees that a network of sufficient capacity is in principle capable of representing the true solution, but says nothing about whether the optimizer actually finds such a configuration. The PINN requirement is also stronger than pointwise approximation of the price surface, since the PDE residual involves first- and second-order partial derivatives; smooth activations are therefore needed so that these derivatives are themselves well behaved. Modern treatments of derivative approximation and deep-learning theory relevant to PDE-solving networks can be found in \parencite{jentzen2023mathintro,grohs2023mathematical,berner2021modern}.

\paragraph{Automatic differentiation.} The partial derivatives appearing in the PDE residual are not estimated by finite differences but computed directly through reverse-mode automatic differentiation (AD) on the computational graph of the network; see \textcite{baydin2018automatic} for a survey and \parencite{paszke2019pytorch} for the PyTorch implementation used in this project. The AD graph must support repeated differentiation so that mixed and higher-order derivatives such as $\partial^2 V/\partial S\partial v$ can be evaluated stably -- another reason for smooth activations. The same machinery makes the loss differentiable with respect to any quantity recorded in the graph, including the Heston parameters $\kappa$, $\theta$, $\xi$, and $\rho$ in the inverse problem, so that the network weights and model parameters can be optimized jointly.

\subsection{Neural-network approximation}

The unknown pricing function is represented by a fully connected feed-forward multilayer perceptron, implemented in PyTorch using the \texttt{torch.nn.Module} interface \parencite{paszke2019pytorch,pytorch_docs}. For the Black--Scholes model, the network maps the two-dimensional input $(S,\tau)$ to a scalar price,
\begin{equation}
    V_{\boldsymbol{w}}(S,\tau) \approx V(S,\tau),
\end{equation}
whereas for the Heston model the input is extended to $(S,v,\tau)$:
\begin{equation}
    V_{\boldsymbol{w}}(S,v,\tau) \approx V(S,v,\tau).
\end{equation}
Here $\boldsymbol{w}$ collects the full set of trainable weights and biases.

A fully connected network with $L$ hidden layers can be written recursively as
\begin{align}
    \mathbf{z}^{(0)} &= \widetilde{\mathbf{x}}, \\
    \mathbf{z}^{(\ell)}
    &=
    \phi\!\left(
        W^{(\ell)}\mathbf{z}^{(\ell-1)}
        +
        \mathbf{b}^{(\ell)}
    \right),
    \qquad \ell=1,\ldots,L, \\
    V_{\boldsymbol{w}}(\mathbf{x})
    &=
    W^{(L+1)}\mathbf{z}^{(L)}
    +
    \mathbf{b}^{(L+1)}.
\end{align}
For Black--Scholes, $\mathbf{x}=(S,\tau)$, while for Heston, $\mathbf{x}=(S,v,\tau)$. Inputs are linearly normalized before being passed through the network; for Heston,
\begin{equation}
    \widetilde{\mathbf{x}}
    =
    \left(
        \frac{S}{S_{\max}},\;
        \frac{v}{v_{\max}},\;
        \frac{\tau}{T_{\max}}
    \right),
\end{equation}
which improves numerical conditioning by keeping the network inputs $\mathcal{O}(1)$ \parencite{glorot2010xavier,goodfellow2016deep}.

\paragraph{Activations.} The activation functions supported by the implementation are $\tanh$, $\mathrm{softplus}(x)=\log(1+e^x)$, GELU \parencite{hendrycks2016gelu}, SiLU/Swish \parencite{ramachandran2017swish}, and the sinusoidal activation of SIREN networks \parencite{sitzmann2020siren}. All of these are at least twice continuously differentiable, which is essential because the PDE residual requires $\partial V/\partial\tau$, $\partial V/\partial S$, $\partial^2 V/\partial S^2$, and -- for Heston -- $\partial^2 V/\partial S\partial v$ and $\partial^2 V/\partial v^2$. Non-smooth activations such as ReLU are unsuited to this setting, since their second derivatives are zero or undefined almost everywhere \parencite{cuomo2022scientific,karniadakis2021physics}. Final activations selected after hyperparameter sweeps were $\tanh$ for the Black--Scholes and forward Heston networks and the sinusoidal SIREN activation for the inverse network.

\paragraph{Weight initialization.} The initialization is matched to the chosen activation, following the implementation in \texttt{activations.py}. Layers with $\tanh$, GELU, or SiLU activations use Xavier normal initialization \parencite{glorot2010xavier}; layers with ReLU or softplus activations use Kaiming (He) normal initialization, which is the appropriate variance scaling for one-sided activations \parencite{goodfellow2016deep}; and layers with sinusoidal activations use the SIREN scheme of \textcite{sitzmann2020siren}, $W_{ij}\sim\mathcal{U}\!\big(-\!\sqrt{6/n_{\mathrm{in}}},\,\sqrt{6/n_{\mathrm{in}}}\big)$, which is essential for stable training of high-frequency sinusoidal networks. Biases are initialized to zero in all cases.

\paragraph{Optimization.} Trainable parameters are optimized by the Adam algorithm \parencite{kingma2015adam} with a cosine-annealing learning-rate schedule \parencite{loshchilov2017sgdr},
\begin{equation}
    \eta_n = \eta_{\min} + \tfrac{1}{2}\!\left(\eta_0 - \eta_{\min}\right)\!\left(1 + \cos\!\frac{n\pi}{T_{\max}}\right),
\end{equation}
with $\eta_{\min} = 10^{-6}$ and $T_{\max}$ equal to the total number of Adam epochs. The base learning rate is $\eta_0 = 5\times10^{-4}$ (Black--Scholes), $5\times10^{-3}$ (forward Heston), or $2\times10^{-3}$ (inverse Heston). To stabilize training under the often ill-conditioned PINN landscape \parencite{wang2021understanding,krishnapriyan2021characterizing}, the global gradient norm is clipped at $1.0$ after each backward pass \parencite{pascanu2013difficulty}:
\begin{equation}
    \boldsymbol{g} \leftarrow \boldsymbol{g}\cdot \min\!\left(1,\;\frac{1}{\|\boldsymbol{g}\|_2}\right),
    \qquad \boldsymbol{g} = \widehat{\nabla_{\boldsymbol{w}}\mathcal{L}}.
\end{equation}
A schematic Adam update then reads
\begin{equation}
    \boldsymbol{w}_{n+1} = \boldsymbol{w}_n - \eta_n\,\boldsymbol{g}_n,
\end{equation}
where $\boldsymbol{g}_n$ is Adam's adaptive moment-based estimate of $\nabla_{\boldsymbol{w}}\mathcal{L}$ after clipping.

\paragraph{Final architectures.} The hyperparameter sweeps reported in the Results section converged on three closely related architectures, summarized in \cref{tab:nn_architectures}. All three are scalar-output multilayer perceptrons trained on PDE, IC, BC, and (for the inverse problem) data losses.

\begin{table}[!htbp]
    \centering
    \footnotesize
    \setlength{\tabcolsep}{4pt}
    \renewcommand{\arraystretch}{1.18}
    \caption{Final network and training configuration for the three PINNs. ``Hidden layers'' counts the internal hidden blocks between the input embedding and the scalar output; widths are constant across layers. Learning rates are base values at the start of the cosine schedule with $\eta_{\min}=10^{-6}$.}
    \label{tab:nn_architectures}
    \begin{tabularx}{\linewidth}{@{} l X X X @{}}
        \toprule
        \textbf{Component} & \textbf{BS PINN} & \textbf{Heston PINN} & \textbf{Inverse PINN} \\
        \midrule
        Input dim                  & $2$ $(S,\tau)$               & $3$ $(S,v,\tau)$             & $3$ $(S,v,\tau)$ \\
        Hidden layers              & $2$                          & $3$                          & $3$ \\
        Width per layer            & $256$                        & $256$                        & $256$ \\
        Activation                 & $\tanh$                      & $\tanh$                      & SIREN (sine) \\
        Weight init                & Xavier normal                & Xavier normal                & SIREN (Sitzmann) \\
        Bias init                  & zero                         & zero                         & zero \\
        Optimizer (Phase 1)        & Adam                         & Adam                         & Adam \\
        Optimizer (Phase 2)        & --                           & --                           & L-BFGS, strong Wolfe \\
        Base LR $\eta_0$           & $5\times10^{-4}$             & $5\times10^{-3}$             & $2\times10^{-3}$ (NN), $10^{-3}$ (Heston) \\
        Min LR $\eta_{\min}$       & $10^{-6}$                    & $10^{-6}$                    & $10^{-6}$ \\
        LR schedule                & cosine-annealing             & cosine-annealing             & cosine-annealing (Phase 1) \\
        Epochs                     & $50{,}000$                   & $50{,}000$                   & $25{,}000$ + $5{,}000$ \\
        Gradient clipping          & $\|g\|_2 \le 1.0$            & $\|g\|_2 \le 1.0$            & Phase 1 only \\
        $\lambda_{\mathrm{pde}}$   & $10$                         & $10$                         & $20$ \\
        $\lambda_{\mathrm{ic}}$    & $5$                          & $10$                         & $10$ \\
        $\lambda_{\mathrm{bc}}$    & $5$                          & $5$                          & $5$ \\
        $\lambda_{\mathrm{data}}$  & --                           & --                           & $1000$ \\
        $\lambda_{\mathrm{Feller}}$ & --                          & --                           & $0.1$ \\
        \bottomrule
    \end{tabularx}
\end{table}

\subsection{Physics-Informed Neural Networks (PINNs)}

A PINN approximates the option-price function by a neural network $V_{\boldsymbol{w}}$ that is trained not only to match payoff and boundary information, but also to satisfy the governing PDE in the interior of the domain \parencite{raissi2019physics,raissi2017physics1}. The total loss for the forward problems is the weighted sum
\begin{equation}\label{eq:pinn_loss}
    \mathcal{L}(\boldsymbol{w})
    =
    \lambda_{\mathrm{pde}}\,\mathcal{L}_{\mathrm{pde}}
    + \lambda_{\mathrm{ic}}\,\mathcal{L}_{\mathrm{ic}}
    + \lambda_{\mathrm{bc}}\,\mathcal{L}_{\mathrm{bc}}.
\end{equation}
The PDE residual loss is the mean squared residual of the pricing equation evaluated at interior collocation points,
\begin{equation}
    \mathcal{L}_{\mathrm{pde}}
    =
    \frac{1}{N_f}\sum_{i=1}^{N_f}
    \left|\mathcal{R}_{\boldsymbol{w}}(\mathbf{x}_i)\right|^2,
\end{equation}
where $\mathbf{x}_i=(S_i,\tau_i)$ for Black--Scholes and $\mathbf{x}_i=(S_i,v_i,\tau_i)$ for Heston, and $\mathcal{R}_{\boldsymbol{w}}$ denotes the strong-form residual obtained by substituting $V_{\boldsymbol{w}}$ and its AD-computed derivatives into \cref{eq:bs_pde} or \cref{eq:heston_pde}. The initial-condition loss enforces the payoff at maturity,
\begin{equation}
    \mathcal{L}_{\mathrm{ic}}
    =
    \frac{1}{N_{\mathrm{ic}}}\sum_{i=1}^{N_{\mathrm{ic}}}
    \left|V_{\boldsymbol{w}}(S_i,0)-\max(S_i-K,0)\right|^2,
\end{equation}
and the boundary-condition loss enforces the prescribed asymptotic behavior,
\begin{equation}
    \mathcal{L}_{\mathrm{bc}}
    =
    \frac{1}{N_{\mathrm{bc}}}\sum_{i=1}^{N_{\mathrm{bc}}}
    \left|V_{\boldsymbol{w}}(\mathbf{x}^{\mathrm{bc}}_i)-g(\mathbf{x}^{\mathrm{bc}}_i)\right|^2,
\end{equation}
with $g$ given by $g(0,\tau)=0$ at the lower asset-price boundary and, where used, $g(S_{\max},\tau)=S_{\max}-Ke^{-r\tau}$ at the upper one. All derivatives entering $\mathcal{R}_{\boldsymbol{w}}$ are computed by reverse-mode AD \parencite{baydin2018automatic,paszke2019pytorch}, so no finite-difference stencil is constructed at any stage.

\paragraph{PINN failure modes and loss balancing.} A recurring difficulty with PINN training is that the gradients of $\mathcal{L}_{\mathrm{pde}}$, $\mathcal{L}_{\mathrm{ic}}$, and $\mathcal{L}_{\mathrm{bc}}$ are typically of very different magnitudes, which produces unbalanced updates and slow convergence. \textcite{wang2021understanding} attribute this to gradient pathologies arising from the multi-objective structure of the loss, \textcite{wang2022expert} analyze the same phenomenon through the lens of the neural tangent kernel, and \textcite{krishnapriyan2021characterizing} document a family of regimes in which naive PINNs fail to converge to the correct PDE solution. In practice, the loss weights $\lambda_{\bullet}$ in \cref{eq:pinn_loss} are treated as hyperparameters chosen by grid search; the values used in this project are reported in \cref{tab:nn_architectures}. General reviews of the PINN literature and surrounding software ecosystem are given by \textcite{karniadakis2021physics}, \textcite{cuomo2022scientific}, and \textcite{lu2021deepxde}; an early application of PINNs to option-pricing PDEs analogous to ours is \textcite{lewis2025pinnsoptions}.

\paragraph{Related deep-learning PDE solvers.} PINNs sit within a broader family of deep-learning PDE solvers. Closely related approaches include the Deep Galerkin Method of \textcite{sirignano2018dgm}, which minimizes a similar residual-based functional over sampled points; the deep BSDE / forward--backward SDE formulation that reframes the pricing problem as a stochastic control problem solved by neural networks \parencite{han2018solving,e2017deep,beck2020deepbsde}; Kolmogorov-PDE methods that learn the solution from Feynman--Kac--simulated paths \parencite{beck2021deep}; and classical Monte Carlo benchmarks \parencite{glasserman2003monte}. Emerging alternative basis architectures such as Kolmogorov--Arnold networks \parencite{liu2024kan} may also prove well suited to PDE tasks but are not pursued here.

\subsection{Inverse Heston calibration}\label{subsec:inverse}

In the inverse problem, the goal is no longer only to approximate the price surface for known parameters: the four Heston parameters $\{\kappa,\theta,\xi,\rho\}$ are themselves treated as trainable variables and recovered from noisy synthetic market observations, jointly with the network weights. The PINN therefore functions as a differentiable calibration engine \parencite{dhiman2023pinns,lewis2025pinnsoptions}, in which AD provides exact gradients of the loss with respect to both $\boldsymbol{w}$ and the model parameters.

\paragraph{Augmented loss.} The training objective is extended by a market-data term and a soft Feller penalty:
\begin{equation}\label{eq:inverse_loss}
    \mathcal{L}_{\mathrm{inv}}(\boldsymbol{w},\kappa,\theta,\xi,\rho)
    =
    \lambda_{\mathrm{pde}}\mathcal{L}_{\mathrm{pde}}
    +\lambda_{\mathrm{ic}}\mathcal{L}_{\mathrm{ic}}
    +\lambda_{\mathrm{bc}}\mathcal{L}_{\mathrm{bc}}
    +\lambda_{\mathrm{data}}\mathcal{L}_{\mathrm{data}}
    +\lambda_{\mathrm{Feller}}\mathcal{L}_{\mathrm{Feller}},
\end{equation}
where
\begin{equation}
    \mathcal{L}_{\mathrm{data}}
    =
    \frac{1}{N_m}\sum_{i=1}^{N_m}
    \left|V_{\boldsymbol{w}}(S_i,v_i,\tau_i) - V_i^{\mathrm{market}}\right|^2.
\end{equation}
The synthetic prices $V_i^{\mathrm{market}}$ are generated from the true Heston parameters $(\kappa^\star,\theta^\star,\xi^\star,\rho^\star)=(2.0,0.04,0.3,-0.7)$ using the semi-analytical formula \eqref{eq:heston_cf_price} and perturbed with additive Gaussian noise of standard deviation $0.5$, which crudely mimics microstructure noise, bid--ask effects, and calibration error. The Feller penalty is implemented as
\begin{equation}\label{eq:feller_loss}
    \mathcal{L}_{\mathrm{Feller}}
    =
    \left[\mathrm{ReLU}\!\left(\xi^2 - 2\kappa\theta\right)\right]^2,
\end{equation}
which is zero whenever the Feller condition $2\kappa\theta\geq\xi^2$ holds \parencite{feller1951two,cox1985cir} and grows quadratically with the magnitude of violation. For the parameters used here, the penalty was numerically zero throughout training and acts only as a safety guard.

\paragraph{Parameter reparameterization.} Calibrating $\kappa$, $\theta$, $\xi$, and $\rho$ directly is awkward because Adam updates are unconstrained and would routinely push these parameters into nonphysical regions ($\kappa\le0$, $|\rho|\ge1$). To eliminate this failure mode the implementation stores \emph{unconstrained} latent variables $\tilde\kappa,\tilde\theta,\tilde\xi,\tilde\rho\in\mathbb{R}$ as the actual trainable parameters and exposes the physical Heston parameters through smooth bijective transforms,
\begin{align}\label{eq:reparam}
    \kappa &= \mathrm{softplus}(\tilde\kappa),\qquad
    \theta = \mathrm{softplus}(\tilde\theta),\qquad
    \xi    = \mathrm{softplus}(\tilde\xi),\\
    \rho   &= \tanh(\tilde\rho),
\end{align}
where $\mathrm{softplus}(x)=\log(1+e^x)>0$ enforces strict positivity and $\tanh$ confines $\rho$ to $(-1,1)$. The optimizer therefore sees the unconstrained $\tilde\bullet$ parameters, while the PDE residual, Feller penalty, and data loss are always evaluated at physically admissible values. User-specified initial guesses $(\kappa_0,\theta_0,\xi_0,\rho_0)=(1.0,0.1,0.5,0.0)$ are mapped to the latent space by the inverse transforms $\tilde\kappa_0=\mathrm{softplus}^{-1}(\kappa_0)$, $\tilde\theta_0=\mathrm{softplus}^{-1}(\theta_0)$, $\tilde\xi_0=\mathrm{softplus}^{-1}(\xi_0)$, $\tilde\rho_0=\mathrm{atanh}(\rho_0)$, with $\rho_0$ clipped into $(-0.999,0.999)$ before the inverse $\tanh$. Gradients with respect to the Heston parameters flow back through the differentiable transforms by chain rule, exactly as for any other AD-tracked quantity.

\paragraph{Two-stage optimization.} The combined parameter space $(\boldsymbol{w};\tilde\kappa,\tilde\theta,\tilde\xi,\tilde\rho)$ has a broad, nonconvex landscape with sharp local minima near the true parameters. A two-stage strategy is used.

\emph{Phase 1 -- Adam (broad basin search).} For the first $25{,}000$ epochs the loss \eqref{eq:inverse_loss} is minimized with Adam \parencite{kingma2015adam} using two parameter groups with separate learning rates: $\eta_0=2\times10^{-3}$ for the NN weights and biases, and $\eta_0=10^{-3}$ for the Heston latent parameters. The smaller learning rate for the latent parameters reflects their direct nonlinear coupling to the PDE residual through the reparameterizations \eqref{eq:reparam}. A single cosine-annealing scheduler is applied to both groups with $\eta_{\min}=10^{-6}$ and $T_{\max}=25{,}000$ \parencite{loshchilov2017sgdr}. Gradient clipping at unit norm is applied after every backward pass for stability \parencite{pascanu2013difficulty}.

\emph{Phase 2 -- L-BFGS (precision fine-tuning).} For the final $5{,}000$ epochs, training switches to the limited-memory BFGS quasi-Newton method \parencite{liu1989limited}, configured with \texttt{max\_iter=1} per outer step, history size $50$, gradient tolerance $10^{-5}$, parameter tolerance $10^{-9}$, and a strong-Wolfe line search. L-BFGS is well suited to the smooth low-curvature regime near a local minimum and routinely reduces the residual by several further orders of magnitude in PINN problems \parencite{raissi2019physics,cuomo2022scientific}. Crucially, gradient clipping is \emph{not} applied during Phase 2: clipping interferes with the curvature estimates and step-length selection of the strong-Wolfe search and is known to break the convergence properties of L-BFGS. Together, the two phases combine Adam's stochastic robustness for the broad search with L-BFGS's deterministic precision near the optimum.

\subsection{Greeks and automatic differentiation}

The Greeks measure the sensitivity of an option price to changes in state variables, time, and model parameters. In classical numerical implementations these quantities are estimated by bump-and-revalue, which becomes prohibitively expensive for mixed, higher-order, or model-specific sensitivities. In the PINN framework the Greeks are obtained directly by differentiating the trained price function $V_{\boldsymbol{w}}$ using the same AD machinery that constructs the PDE residual \parencite{baydin2018automatic,paszke2019pytorch}. First-order, second-order, and third-order derivatives can therefore be extracted from the learned surface without any finite-difference stencil or re-evaluation of the pricer.

\paragraph{Black--Scholes Greeks: which are accessible by AD?} A subtle but important point is that the Black--Scholes PINN takes only $(S,\tau)$ as inputs; the volatility $\sigma$, risk-free rate $r$, and dividend yield $q$ are baked into the trained weights as constants. Consequently, only Greeks expressible as derivatives with respect to $S$ or $\tau$ are directly available from the trained network -- namely Delta, Theta, Gamma, Charm, Speed, and Color. Volatility- and rate-sensitive Greeks (Vega $\partial V/\partial\sigma$, Vomma, Vanna, Zomma, $\rho_r=\partial V/\partial r$, $\Phi=\partial V/\partial q$) are \emph{not} obtainable from this network by AD; recovering them would require training a \emph{parametric PINN} in which $\sigma$, $r$, and $q$ enter as additional inputs. In the Heston setting the variance state $v$ is already an input, so the variance-direction sensitivities Vega$_v$, Volga$_v$, Vanna$_v$, and Zomma$_v$ \emph{are} directly accessible by AD, in addition to the $S$- and $\tau$-Greeks. The Greeks computed in this project are summarized in \cref{tab:greeks_expressions}, with the BS-only volatility/rate Greeks listed for reference and marked as unavailable from the present non-parametric network.

\paragraph{$\sigma$-Greeks versus $v$-Greeks.} In the Black--Scholes model, volatility is a constant model parameter $\sigma$, so Vega-type quantities are derivatives with respect to $\sigma$. In the Heston model, volatility enters through the variance state variable $v$, and the corresponding quantities measure how the learned price surface changes along the variance dimension. The Heston quantities Vega$_v$, Volga$_v$, Vanna$_v$, and Zomma$_v$ should therefore be interpreted as variance-state sensitivities rather than direct Black--Scholes volatility sensitivities.

\begin{table}[!htbp]
    \centering
    \small
    \setlength{\tabcolsep}{4pt}
    \renewcommand{\arraystretch}{1.18}
    \caption{Greeks computed from the trained PINN price function. Black--Scholes sensitivities are written with respect to the constant volatility parameter $\sigma$; Heston sensitivities are written with respect to the variance state $v$. Rows marked ``BS$^\star$'' require a parametric PINN with $\sigma,r,q$ as inputs and are not directly accessible from the non-parametric network used here.}
    \label{tab:greeks_expressions}
    \begin{tabularx}{\linewidth}{@{} l X c c @{}}
        \toprule
        \textbf{Greek} & \textbf{Expression} & \textbf{Order} & \textbf{Model} \\
        \midrule

        \multicolumn{4}{@{}l}{\textbf{Price level}} \\
        Price          & $V$                                  & --   & BS/HS \\

        \addlinespace[0.25em]
        \multicolumn{4}{@{}l}{\textbf{First-order sensitivities}} \\
        Delta $(\Delta)$ & $\partial V / \partial S$           & 1st  & BS/HS \\
        Theta $(\Theta)$ & $\partial V / \partial \tau$        & 1st  & BS/HS \\
        Rho $(\rho_r)$   & $\partial V / \partial r$           & 1st  & BS$^\star$ \\
        Phi $(\Phi)$     & $\partial V / \partial q$           & 1st  & BS$^\star$ \\
        Vega             & $\partial V / \partial \sigma$      & 1st  & BS$^\star$ \\
        Vega$_v$         & $\partial V / \partial v$           & 1st  & HS \\

        \addlinespace[0.25em]
        \multicolumn{4}{@{}l}{\textbf{Second-order sensitivities}} \\
        Gamma $(\Gamma)$ & $\partial^2 V / \partial S^2$       & 2nd  & BS/HS \\
        Vomma            & $\partial^2 V / \partial \sigma^2$  & 2nd  & BS$^\star$ \\
        Volga$_v$        & $\partial^2 V / \partial v^2$       & 2nd  & HS \\
        Vanna            & $\partial^2 V / \partial S \partial \sigma$ & 2nd & BS$^\star$ \\
        Vanna$_v$        & $\partial^2 V / \partial S \partial v$      & 2nd & HS \\
        Charm            & $\partial^2 V / \partial S \partial \tau$   & 2nd & BS/HS \\

        \addlinespace[0.25em]
        \multicolumn{4}{@{}l}{\textbf{Third-order sensitivities}} \\
        Speed            & $\partial^3 V / \partial S^3$       & 3rd  & BS/HS \\
        Zomma            & $\partial^3 V / \partial S^2 \partial \sigma$ & 3rd & BS$^\star$ \\
        Zomma$_v$        & $\partial^3 V / \partial S^2 \partial v$     & 3rd & HS \\
        Color            & $\partial^3 V / \partial S^2 \partial \tau$  & 3rd & BS/HS \\

        \addlinespace[0.25em]
        \multicolumn{4}{@{}l}{\textbf{PDE diagnostic}} \\
        PDE residual     & $\mathcal{R}_{\boldsymbol{w}}$      & --   & BS/HS \\

        \bottomrule
    \end{tabularx}
\end{table}

\subsection{Collocation data and sampling strategy}\label{subsec:collocation}

PINNs are mesh-free, but they are not data-free: their performance depends strongly on where collocation points are sampled \parencite{lu2021deepxde,wang2022expert}. The collocation design therefore determines which regions of the pricing domain are emphasized during training. Since option-price surfaces contain the strongest curvature near the strike and near expiry -- where the kinked payoff $\max(S-K,0)$ propagates a sharp feature backward in $\tau$ -- purely uniform sampling underrepresents precisely the regions that most challenge the network. We use an explicit mixture distribution that combines a uniform background with concentrations near the strike, near expiry, and (for Heston) near the initial variance level.

\subsubsection{Black--Scholes forward problem}

For the Black--Scholes forward problem, the interior collocation set of size $N_f=5000$ is drawn from a three-component mixture:
\begin{itemize}
    \item \textbf{60\% uniform background} on the rectangle $[0,S_{\max}]\times[0,T_{\max}]$;
    \item \textbf{20\% near the strike}, with $S\sim\mathcal{N}(K,(0.2K)^2)$ clipped to $[0,S_{\max}]$ and $\tau$ uniform on $[0,T_{\max}]$;
    \item \textbf{20\% near expiry}, with $\tau\sim\mathrm{Beta}(0.5,2.0)\cdot T_{\max}$ -- a heavily left-skewed distribution that concentrates samples near the payoff singularity at $\tau=0$ where the curvature is highest -- and $S$ uniform.
\end{itemize}
The initial-condition set of size $N_{\mathrm{ic}}=1000$ is a 50/50 mixture of uniform $S$ and $S\sim\mathcal{N}(K,(0.15K)^2)$ clipped to $[0,S_{\max}]$, all at $\tau=0$. Boundary points are split evenly between the lower boundary $S=0$ (where $V=0$ is enforced) and the upper boundary $S=S_{\max}$, each with $\tau$ uniform on $[0,T_{\max}]$. The final BS PINN imposes only the lower-boundary penalty $V(0,\tau)=0$ in $\mathcal{L}_{\mathrm{bc}}$; the upper-boundary points are generated but optional. \Cref{tab:bs_collocation_summary} summarizes the dataset.

\begin{table}[!htbp]
    \centering
    \small
    \setlength{\tabcolsep}{6pt}
    \renewcommand{\arraystretch}{1.18}
    \caption{Black--Scholes collocation set summary.}
    \label{tab:bs_collocation_summary}
    \begin{tabularx}{\linewidth}{@{} X r @{}}
        \toprule
        \textbf{Quantity} & \textbf{Value} \\
        \midrule

        \multicolumn{2}{@{}l}{\textbf{Collocation points}} \\
        Interior points $N_f$ (60/20/20 mixture)                & $5000$ \\
        Initial-condition points $N_{\mathrm{ic}}$ (50/50 mix)  & $1000$ \\
        Boundary $S=0$                                          & $250$ \\
        Boundary $S=S_{\max}$                                   & $250$ \\

        \addlinespace[0.25em]
        \multicolumn{2}{@{}l}{\textbf{Domain}} \\
        $S$ range & $[0.00,\ 300.00]$ \\
        $\tau$ range & $[0.00,\ 1.00]$ \\

        \bottomrule
    \end{tabularx}
\end{table}

\subsubsection{Heston forward problem}

The Heston forward problem extends the input domain from $(S,\tau)$ to $(S,v,\tau)$ and introduces the mixed-derivative term $\partial^2 V/\partial S\partial v$. To maintain coverage of the enlarged domain, the interior set is doubled to $N_f=10{,}000$ and drawn from a four-component mixture:
\begin{itemize}
    \item \textbf{40\% uniform background} on $[0,S_{\max}]\times[v_{\min},v_{\max}]\times[0,T_{\max}]$;
    \item \textbf{20\% near the strike}, with $S\sim\mathcal{N}(K,(0.2K)^2)$ clipped to $[0,S_{\max}]$;
    \item \textbf{20\% near expiry}, with $\tau\sim\mathrm{Beta}(0.5,2.0)\cdot T_{\max}$;
    \item \textbf{20\% near the initial variance}, with $v\sim\mathcal{N}(v_0,(0.3 v_0)^2)$ clipped to $[v_{\min},v_{\max}]$.
\end{itemize}
The initial-condition set of size $N_{\mathrm{ic}}=2000$ is again a 50/50 mixture, this time combining uniform $(S,v)$ with samples clustered near the strike in $S$ and near $v_0$ in $v$. The collocation file contains \emph{four} boundary sets of $250$ points each: $S=0$, $S=S_{\max}$, $v=v_{\min}$, and $v=v_{\max}$. However, the final forward Heston PINN imposes only the two asset-price boundaries in $\mathcal{L}_{\mathrm{bc}}$ (with $V(0,v,\tau)=0$ at the lower boundary), because the limiting behavior of $V$ in the variance direction is not a simple zero-value Dirichlet condition. The variance-boundary points are loaded but excluded from the boundary loss; this choice is consistent with the no-arbitrage requirement at $v\to 0$ and the lack of an elementary expression at $v=v_{\max}$.

\begin{table}[!htbp]
    \centering
    \small
    \setlength{\tabcolsep}{6pt}
    \renewcommand{\arraystretch}{1.18}
    \caption{Heston collocation set summary. The generated dataset contains four boundary sets, but the final Heston PINN imposes boundary penalties only on the two asset-price boundaries.}
    \label{tab:hs_collocation_summary}
    \begin{tabularx}{\linewidth}{@{} X r @{}}
        \toprule
        \textbf{Quantity} & \textbf{Value} \\
        \midrule

        \multicolumn{2}{@{}l}{\textbf{Collocation points used in final training}} \\
        Interior points $N_f$ (40/20/20/20 mixture)             & $10{,}000$ \\
        Initial-condition points $N_{\mathrm{ic}}$ (50/50 mix)  & $2000$ \\
        Boundary $S=0$                                          & $250$ \\
        Boundary $S=S_{\max}$                                   & $250$ \\

        \addlinespace[0.25em]
        \multicolumn{2}{@{}l}{\textbf{Generated but not imposed in final boundary loss}} \\
        Boundary $v=v_{\min}$                                   & $250$ \\
        Boundary $v=v_{\max}$                                   & $250$ \\

        \addlinespace[0.25em]
        \multicolumn{2}{@{}l}{\textbf{Domain}} \\
        $S$ range & $[0.00,\ 300.00]$ \\
        $v$ range & $[0.01,\ 1.00]$ \\
        $\tau$ range & $[0.00,\ 1.00]$ \\

        \bottomrule
    \end{tabularx}
\end{table}

\subsubsection{Heston inverse problem}

For the inverse calibration problem the training set combines the same physics collocation distribution as the forward Heston PINN with an additional $N_m=200$ noisy market observations. The market points are generated by sampling $(S,v,\tau)$ from the interior distribution, evaluating the semi-analytical Heston call price \eqref{eq:heston_cf_price} at the true parameters, and adding i.i.d.\ Gaussian noise of standard deviation $0.5$ to the resulting prices. The variance boundaries are again \emph{not} imposed in the boundary loss, and in fact for the inverse model only the single lower asset-price boundary $S=0$ enters $\mathcal{L}_{\mathrm{bc}}$, the rationale being that the calibration signal at $S=S_{\max}$ is implicitly carried by both the PDE residual and the market-price observations. \Cref{tab:inverse_collocation_summary} summarizes the dataset.

\begin{table}[!htbp]
    \centering
    \small
    \setlength{\tabcolsep}{6pt}
    \renewcommand{\arraystretch}{1.18}
    \caption{Inverse-problem dataset summary. The boundary points at $S=S_{\max}$ and at $v=v_{\min}, v=v_{\max}$ are generated but not imposed in the final inverse boundary loss.}
    \label{tab:inverse_collocation_summary}
    \begin{tabularx}{\linewidth}{@{} X r @{}}
        \toprule
        \textbf{Quantity} & \textbf{Value} \\
        \midrule

        \multicolumn{2}{@{}l}{\textbf{Collocation points}} \\
        Interior points $N_f$               & $10{,}000$ \\
        Initial-condition points $N_{\mathrm{ic}}$ & $2000$ \\
        Boundary $S=0$ (imposed)            & $250$ \\
        Boundary $S=S_{\max}$ (generated)   & $250$ \\
        Boundary $v=v_{\min}$ (generated)   & $250$ \\
        Boundary $v=v_{\max}$ (generated)   & $250$ \\
        Market-data points $N_m$            & $200$ \\

        \addlinespace[0.25em]
        \multicolumn{2}{@{}l}{\textbf{Collocation domain}} \\
        $S$ range   & $[0.00,\ 300.00]$ \\
        $v$ range   & $[0.01,\ 1.00]$ \\
        $\tau$ range & $[0.00,\ 1.00]$ \\

        \addlinespace[0.25em]
        \multicolumn{2}{@{}l}{\textbf{Market-data observation domain}} \\
        Market $S$ range   & $[51.32,\ 249.55]$ \\
        Market $v$ range   & $[0.0111,\ 0.4990]$ \\
        Market $\tau$ range & $[0.0529,\ 0.9970]$ \\
        Market $V$ range   & $[0.0000,\ 152.8329]$ \\
        Noise std. $\sigma_{\mathrm{noise}}$ & $0.5$ \\

        \bottomrule
    \end{tabularx}
\end{table}

\subsection{Computational environment and tooling}

All experiments were implemented in Python using PyTorch \parencite{paszke2019pytorch,pytorch_docs}, with reverse-mode AD as the only differentiation primitive. Numerical experiments were executed on cloud-hosted GPU runtimes through Google Colab \parencite{google_colab_faq}, primarily on NVIDIA A100 \parencite{nvidia_a100} and H100 \parencite{nvidia_h100} Tensor Core accelerators. GPU parallelism is particularly valuable for PINNs because each training step evaluates the network and several of its derivatives at thousands of collocation points; the resulting batched dense linear algebra and AD passes map naturally onto Tensor Core hardware. Local development and debugging were performed on a personal workstation in VS Code \parencite{vscode}, with AI-assisted coding tools \parencite{cursor,google_antigravity,anthropic_claude_code} used in a supporting role for code navigation, refactoring, and LaTeX formatting. These tools were not treated as authoritative mathematical sources; the model selection, numerical experiments, interpretation of results, and final responsibility for the manuscript remain with the author.
